---
title: "Homework 5"
author: "Evan Hall"
date: "02/05/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = "~/Lab5")
```

# Load packages
```{r Load packages, message=FALSE, warning=FALSE, include=FALSE, results='hide'}
library(tidyverse)
library(readxl)
library(broom)
library(cowplot)
set.seed(7)
```

Save any data frames generated to curated_data directory, save all plots to figures directory. Based on assumptions, clearly state which statistical test will be used and why. Include any statistics (p- or R-values) as comments after the tests.

# Question 1

Use the SCFA wide dataframe from Lab 5. Check assumptions, conduct statistical tests, make a plot, to answer the question, "Which supplement causes the greatest increase in acetate concentrations?"
```{r}
# import and filter data set (q1_df)
q1_df <- read_delim(file = "raw_data/scfa_wide.txt", 
                        delim = "\t", escape_double = FALSE, 
                        trim_ws = TRUE, na=c("NA"),
                        col_names = TRUE,
                        col_types = cols(participant_id = col_character(),
                                         semester = col_character(),
                                         supplement_consumed = col_character(),
                                         quantity_compliant = col_character(),
                                         frequency = col_character(),
                                         acetate_delta = col_double(),
                                         butyrate_delta = col_double(),
                                         propionate_delta = col_double())) %>%
  filter(semester != "Winter2015",
         quantity_compliant != "no")
```

```{r}
# check assumptions
# Variance
bartlett.test(acetate_delta ~ supplement_consumed, data = q1_df)
# ALTERNATIVE 
```

```{r}
# conduct statistical test
aov_results <- aov(acetate_delta ~ supplement_consumed, data = q1_df)
summary(aov_results)
```
```{r}
tuk_delta <- tidy(TukeyHSD(aov_results)) %>%
  mutate(adj.p.value = round(adj.p.value, digits = 2)) #round resulting p-values
 #filter Tukey output for significant results 
write_delim(tuk_delta, path = "curated_data/tukey_result_acetate_delta.txt",
            delim = "\t", col_names = TRUE, quote = FALSE)
```

```{r}
# make and save plot (q1_plot)
q1_plot <-  q1_df %>%
  filter(semester != "Winter2015",
         quantity_compliant != "no")
# plot
q1_df %>%
  ggplot(aes(x = supplement_consumed, 
             y = acetate_delta, 
             color = supplement_consumed)) + 
  geom_hline(yintercept = 0, linetype = "dashed") + 
  geom_violin() + 
  geom_jitter() + 
  xlab(NULL) + 
  ylab("Delta Acetate Concentration") + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

# run the test
kruskal.test(supplement_consumed ~ acetate_delta, q1_df)

write_delim(q1_df, path = "figures/q1.txt",
            delim = "\t", col_names = TRUE, quote = FALSE)
```

What are your conclusions regarding the reseach question? 
Because the p-value is extremely high .5005 at the .05 alpha level, then we would fail to reject the null hypothesis. There is no difference any of the variables that influence the change in acetate concentration. 
<type answer here>


# Question 2

Use the SCFA wide dataframe from Lab 5. Check assumptions, conduct statistical tests, make a plot, to answer the question, "Which supplement causes the greatest increase in propionate concentrations?"
```{r}
# import and filter data set (q2_df)

q2_df <- read_delim(file = "raw_data/scfa_wide.txt", 
                        delim = "\t", escape_double = FALSE, 
                        trim_ws = TRUE, na=c("NA"),
                        col_names = TRUE,
                        col_types = cols(participant_id = col_character(),
                                         semester = col_character(),
                                         supplement_consumed = col_character(),
                                         quantity_compliant = col_character(),
                                         frequency = col_character(),
                                         acetate_delta = col_double(),
                                         butyrate_delta = col_double(),
                                         propionate_delta = col_double())) %>%
  filter(semester != "Winter2015",
         quantity_compliant != "no")
```

```{r}
# check assumptions
bartlett.test(propionate_delta ~ supplement_consumed, data = q1_df)
# Variances are equal
# Normality
s1 <- q2_df %>%
  filter(supplement_consumed == "Accessible") 
shapiro.test(s1$propionate_delta) #p-value = .23 NORMAL

s2 <- q2_df %>%
  filter(supplement_consumed == "Arabino") 
shapiro.test(s2$propionate_delta) #p-value = .97 NORMAL 

s3 <- q2_df %>%
  filter(supplement_consumed == "BRMPS+Accessible")
shapiro.test(s3$propionate_delta) #p-value = .09 NOT NORMAL

s4 <- q2_df %>%
  filter(supplement_consumed == "HiMaize") 
shapiro.test(s4$propionate_delta) #p-value = .02 NOT NORMAL

s5 <- q2_df %>%
  filter(supplement_consumed == "HiMaize+BRMPS") 
shapiro.test(s5$propionate_delta) #p-value = .015 NOT NORMAL

s6 <- q2_df %>%
  filter(supplement_consumed == "Inulin") 
shapiro.test(s6$propionate_delta) #p-value = .009 NOT NORMAL

s7 <- q2_df %>%
  filter(supplement_consumed == "BRMPS") 
shapiro.test(s7$propionate_delta) #p-value = 7.829e-05 NOT NORMAL

s8 <- q2_df %>%
  filter(supplement_consumed == "LOODAT") 
shapiro.test(s8$propionate_delta) #p-value = .89 NORMAL 

s9 <- q2_df %>%
  filter(supplement_consumed == "none")
shapiro.test(s9$propionate_delta) #p-value = .0099 NOT NORMAL

s10 <- q2_df %>%
  filter(supplement_consumed == "Psyllium") 
shapiro.test(s10$propionate_delta) #p-value = .5866 NORMAL

s11 <- q2_df %>%
  filter(supplement_consumed == "Psyllium+BRMPS") 
shapiro.test(s11$propionate_delta) #p-value = .03922 NOT NORMAL

s12 <- q2_df %>%
  filter(supplement_consumed == "transition_HiMaize") 
shapiro.test(s12$propionate_delta) #p-value = .7647 NORMAL

# sample size
q2_df %>%
  group_by(supplement_consumed) %>%
  summarise(sample_size = n())
```

```{r}
# conduct statistical test
kruskal.test(supplement_consumed ~ propionate_delta, q2_df)
```

```{r}
# make and save plot (q2_plot)
q2_plot <-  q2_df %>%
  filter(semester != "Winter2015",
         quantity_compliant != "no")
# plot
q2_df %>%
  ggplot(aes(x = supplement_consumed, 
             y = propionate_delta, 
             color = supplement_consumed)) + 
  geom_hline(yintercept = 0, linetype = "dashed") + 
  geom_violin() + 
  geom_jitter() + 
  xlab(NULL) + 
  ylab("Delta Propionate Concentration") + 
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))

write_delim(q2_plot, path = "figures/q2_plot.txt",
            delim = "\t", col_names = TRUE, quote = FALSE)
```

What are your conclusions regarding the reseach question?
Because the p-value is greater than the .05 significance level (.5637), we are going to fail to reject the null hypothesis. There is no difference any of the variables that influence the change in propionate concentration. 
<type answer here>


# Question 3

Use the SCFA wide dataframe from Lab 5. Subset and filter the data frame, check assumptions, and choose the appropriate statistical tests, and create a plot to answer the following research question: "Does the proportion of acetate reponses differ between individuals who consumed BRMPS potato (BRMPS) or LOODAT potato supplements?"

```{r}
# import and filter data set (q3_df)
q3_df <- q1_df %>%
  filter(supplement_consumed == "LOODAT" | supplement_consumed == "BRMPS",
         quantity_compliant != "no",
         semester != "Winter2015") %>%
  select(-butyrate_delta, -propionate_delta) %>% 
  add_column(ace_status = "-") %>%
  drop_na()
  
```

```{r}
# check assumptions
#sample size
q3_df %>%
  group_by(supplement_consumed) %>%
  summarise(sample_size = n())
# BRMPS is larger than 30 
q3_df$ace_status[q3_df$acetate_delta > 0] <- "positive"
q3_df$ace_status[q3_df$acetate_delta < 0] <- "negative"

# normality
s13 <- q3_df %>%
  filter(supplement_consumed == "BRMPS") 
shapiro.test(s13$acetate_delta) #p-value = .04394 NOT NORMAL

s14 <- q3_df %>%
  filter(supplement_consumed == "LOODAT") 
shapiro.test(s14$acetate_delta) #p-value = .1042 NORMAL
```
```{r}
ace_tab <- with(q3_df, table(supplement_consumed, ace_status))
```

```{r}
q3_df %>%
  group_by(supplement_consumed, ace_status) %>%
  summarise(Counts = n())
```

```{r}
# conduct statistical test
prop.test(ace_tab, 
          alternative = "two.sided",
          correct = FALSE)
```

```{r}
# make and save plot (q3_plot)
```

| | Increased acetate | Total participants | Percent increase |
|:-----:|:-----:|:-----:|:-----:|
| BRMPS | 75 | 128 | 58.6% |
| LOODAT | 9 | 16 | 56.25% | 

What are your conclusions regarding the reseach question?
Because the p-value .8577 is greater than the .05 significance level, we are going to fail to reject the null hypothesis. There is no difference in the responses of BRMPS or LOODAT based on the proportion of acetate.

<type answer here>


# Question 4

Use the SCFA wide dataframe from Lab 5. Subset and filter the data frame, check assumptions, and choose the appropriate statistical tests, and create a plot to answer the following research question: "Does the proportion of propionate reponses differ between individuals who consumed BRMPS potato (BRMPS) or LOODAT potato supplements?"

```{r}
# import and filter data set (q4_df)
q4_df <- q1_df %>%
  filter(supplement_consumed == "LOODAT" | supplement_consumed == "BRMPS",
         quantity_compliant != "no",
         semester != "Winter2015") %>%
  select(-butyrate_delta, -acetate_delta) %>% 
  add_column(prop_status = "-") %>%
  drop_na()
```

```{r}
# check assumptions
#sample size
q4_df %>%
  group_by(supplement_consumed) %>%
  summarise(sample_size = n())
# BRMPS is a large enough sample size > 30
q4_df$prop_status[q4_df$propionate_delta > 0] <- "positive"
q4_df$prop_status[q4_df$propionate_delta < 0] <- "negative"

# normality
s15 <- q4_df %>%
  filter(supplement_consumed == "BRMPS") 
shapiro.test(s15$propionate_delta) #p-value = 7.829e-05 NOT NORMAL

s16 <- q4_df %>%
  filter(supplement_consumed == "LOODAT") 
shapiro.test(s16$propionate_delta) #p-value = 0.8912 NORMAL
```

```{r}
prop_tab <- with(q4_df, table(supplement_consumed, prop_status))

q4_df %>%
  group_by(supplement_consumed, prop_status) %>%
  summarise(Counts = n())
```

```{r}
# conduct statistical test
prop.test(prop_tab, 
          alternative = "two.sided",
          correct = FALSE)
```

```{r}
# make and save plot (q4_plot)
```

| | Increased propionate | Total participants | Percent increase |
|:-----:|:-----:|:-----:|:-----:|
| BRMPS | 54 | 127 | 42.5% |
| LOODAT | 7 | 16 | 43.75% | 

What are your conclusions regarding the reseach question?
Because the p-value .9253 is greater than the .05 significance level, we are going to fail to reject the null hypothesis. There is no difference in the responses of BRMPS or LOODAT based on the proportion of propionate.

<type answer here>


-----
end